Binning Strategies and Related Loss for Large Data

### 2015-03-31 Paper.Rnw
more proofreading

Binning data for the purpose of creating a binned scatterplot requires a choice of algorithm as well as a choice of parameters associated with the  binning algorithm.  This section aims to compare binning algorithms and identify the best parameter choices for minimizing loss under a number of scenarios.  \ktm Some results are derived \hh{XXX check the sentence} optimal choices through analytical proofs while others are recommendation based on numerical from empirical exploration of loss from binning.  }

Whether analytical or empirical, data is needed to demonstrate how loss is impacted by binning choices. Datasets were simulated from bivariate distributions to be used throughout this section: Exponential, Normal and Uniform.  These distributions were selected for their variety in terms of shape, center and spread. 100,000 observation pairs were simulated for each set of simulated data, and the $x$ and $y$ values were drawn independently from univariate distributions; Exponental(11), Normal(50,$11^2$) and Uniform(0,100), respectively.  The parameters were selected to have data values roughly span the region $[0,100]^2$. The simulated data can be found in the top row of Figure~\ref{fig:SimDataScatterplots}.

@@

\ktm{ These simulated datasets are from continous distributions, and thus the values are recorded to many decimal places. Real data is recorded to only the number of digits that the measurements allow; in many cases rounded to the nearest whole dollar or five minutes or two miles. To observe binning loss under more realistic conditions, three corresponding data sets were also created by rounding the values from the original simulated data sets to the nearest even number. This coarse version of the original simulated data is displayed in the bottom row of Figure~\ref{fig:SimDataScatterplots}. By exploring the loss properties for both the \textit{fine} and \textit{coarse} versions of the data, we identify which binning options are robust to the resolution to which data is recorded. }

@@

For rectangular binning there are many specification options; including type of algorithm, location of the origin and binwidths for each dimension. To explore the spatial loss properties under different binning approaches we begin with a comparison of standard and random binning. If equivalent bin widths are used, the net spatial loss under standard binning will always be less than or equal to the net spatial loss under random binning. This is because the minimal spatial loss for each data point under random binning is to allocate to the nearest bin center, which is how the point would be allocated in standard binning.

Figure~\ref{fig:SpatialLossCompare} displays the proportion of potential spatial loss from binning each set of simulated data with standard and random binning algorithms using square bins with a sizes ranging from 2 units$^2$ to 20 units$^2$. It is clear to see that the spatial loss for random binning is always higher than for standard binning and that the loss grows as binwidth increases. The proportion of potential spatial loss from random binning increases with binwidth at a faster rate than with standard binning, as indicated by the widening gap and steeper slopes.

@@

All rectangular binning algorithms require the specification of a bin width for each dimension and a binning origin. Smaller binwidths coincide with smaller spatial losses, but reducing bin sizes comes with a few important caveats. With smaller bins comes more computation requirements; a non-negligable consideration in settings with truly massive data. Additionally, the binwidth should be an integer multiple of the resolution of the data to avoid undesirable visual attributes in the binned scatterplot.

@@

\ktm{
\begin{itemize}
\item Binning Algorithm: total and net spatial losses are always larger for random binning than standard (analytical proof)
\item Origin: 
  \begin{itemize}
    \item Doesn't much matter for fine resolution data but it does for coarse data.
    \item If recorded to resolution $\alpha$, origin at $x_{(1)} - \alpha/2$ is optimal (analytical proof then demo) Proof in Appendix~\ref{proof:offset}.
  \end{itemize} 
\item Binwidth: 
  \begin{itemize}
    \item generally smaller $\omega \rightarrow$ smaller $L^s$. 
    \item for fine data any binwidth is fine
    \item for coarse data need binwidths $\omega \ge \alpha$ (show plot with whitespace gaps)
    \item for coarse data binwidths should be scalar multiple of $\alpha$ (show plot with strips)
    \item Does the distribution type matter? 
  \end{itemize}
\end{itemize}
}

@@

Investigation of the loss sources we might be tempted to conclude that the standard binning algorithm is superior due to lower loss, however the random binning algorithm displays strong advantages when encountering problematic data structure issues. Random binning is especially advantageous when discrete data values are recorded at a consistent increments but we bin our data using a binwidth that is a non-integer multiple of the incremental widths. In this scenario, we see that the non-synchronous data increments and bin widths will cause bins to contain inconstant numbers of possible data value. Assuming that data values are spread uniformly over the discrete values we want a visualization that displays uniform frequency. Using the standard binning algorithm the misalignment of data increment and bin widths will lead to and artificial frequency discrepancy that manifests in a striped pattern in the visualization. The random binning algorithm is expected to display the intended uniform frequency pattern because the randomization of data values to bin centers mitigates the problematic bin width selection.

@@

Another thing to note about Figure ~\ref{fig:LossComparison1} is that under the standard binning algorithm the numerical loss is a monotone increasing function of the bin width, but visual loss is not a strictly increasing function of bin width.  This is apparent in the order of colored lines based on the Strike Out binning size.  This is because the numerical loss function sums loss over all observations, which means the distance from the points to the numerical center of bins can only increase as binwidth increases. Whereas the visual loss function sums loss over bin, thus as the binwidth increases the number of bins decreases which leads to the potential that the overall visual loss may decrease even though the distance from numerical centers to visual centers may increase for individual bins.

@@

We may also use this example to highlight the advantages of random binning. The two binned scatterplots at the top of Figure~\ref{fig:binning-comparison} show bins of the same width, but clear differences between the binning results are evident. The standard binning algorithm has rounding artifacts that are clearly visible in the vertical strips. These vertical stripes are from the computational rounding issues encountered if we use the standard binning when we have the data values were recorded on an evenly spaces discrete scale (in our case, integers) and the binwidths are integer multiples of the data increments.  This reinforces the argument that in practice the random algorithm can produce a graph that better represents the true shape of the data than the standard algorithm even if the numerical and visual loss is higher.

\textbf{Are we sure that this is because of an odd binwidths and not just that the bins were not centered on an integer?} In contrast, when the bin widths are odd, the two binning algorithms yield nearly indistinguishable binned scatterplots, as shown in the bottom row of Figure~\ref{fig:binning-comparison}.\textbf{I think that strips disappearing in (5,25) bins is because of where we START the binning not that the bin widths are even}

>>>>

Binning data for the purpose of creating a binned scatterplot requires a choice of algorithm as well as a choice of parameters associated with the  binning algorithm.  This section aims to compare binning algorithms and identify the best parameter choices for minimizing loss under a number of scenarios.  \ktm{Some results are derived \hh{XXX check the sentence} optimal choices through analytical proofs while others are recommendation based on numerical from empirical exploration of loss from binning.  }

Whether analytical or empirical, data is needed to demonstrate how loss is impacted by binning choices.

\hh{Just list the three different scenarios:
\begin{itemize}
\item {\bf Scenario I:} $x \sim \text{Exp}_\lambda, y \sim \text{Exp}_\lambda$ with $\lambda = 11$
\end{itemize}}
Datasets were simulated from bivariate distributions to be used throughout this section: Exponential, Normal and Uniform.  These distributions were selected for their variety in terms of shape, center and spread. 100,000 observation pairs were simulated for each set of simulated data, and the $x$ and $y$ values were drawn independently from univariate distributions; Exponental(11), Normal(50,$11^2$) and Uniform(0,100), respectively.  The parameters were selected to have data values roughly span the region $[0,100]^2$. The simulated data can be found in the top row of Figure~\ref{fig:SimDataScatterplots}.

@@

These simulated datasets are from continous distributions, and thus the values are recorded to many decimal places. Real data is recorded to only the number of digits that the measurements allow; in many cases rounded to the nearest whole dollar or five minutes or two miles. To observe binning loss under more realistic conditions, we created three  data sets by rounding the values from the originally simulated data sets to the nearest even number. This coarse version of the original simulated data is displayed in the bottom row of Figure~\ref{fig:SimDataScatterplots}. By exploring the loss properties for both the \textit{fine} and \textit{coarse} versions of the data, we identify which binning options are robust to the resolution to which data is recorded.

@@

For rectangular binning there are many specification options; including type of algorithm, location of the origin and binwidths for each dimension. To explore the spatial loss properties under different binning approaches we begin with a comparison of standard and random binning: For equal bin widths, the net spatial loss under standard binning is always less than or equal to the net spatial loss under random binning. This is because the minimal spatial loss for each data point under random binning is to allocate to the nearest bin center, which is how the point would be allocated in standard binning.

Figure~\ref{fig:SpatialLossCompare} displays the proportion of potential spatial loss from binning each set of simulated data with standard and random binning algorithms using square bins with a sizes ranging from 2 units$^2$ to 20 units$^2$.
Obviously, the spatial loss for random binning is always higher than for standard binning and the loss grows as bin width increases. The proportion of potential spatial loss from random binning increases with bin width at a faster rate than from standard binning, as indicated by the widening gap between the lines and the steeper slopes.

@@

All rectangular binning algorithms require the specification of a bin width for each dimension and a binning origin. Smaller bin widths coincide with smaller spatial losses, but reducing bin sizes comes with a few important caveats. With smaller bins comes more computation requirements; a non-negligible consideration in settings with truly massive data. Additionally, the bin width should be an integer multiple of the resolution of the data to avoid undesirable visual attributes in the binned scatterplot.

@@

\ktm{
\begin{itemize}
\item Binning Algorithm: total and net spatial losses are always larger for random binning than standard (analytic proof)
\item Origin: 
  \begin{itemize}
    \item Doesn't much matter for fine resolution data but it does for coarse data.
    \item If recorded to resolution $\alpha$, origin at $x_{(1)} - \alpha/2$ is optimal (analytically proof then demo) Proof in Appendix~\ref{proof:offset}.
  \end{itemize} 
\item Bin width: 
  \begin{itemize}
    \item generally smaller $\omega \rightarrow$ smaller $L^s$. 
    \item for fine data any bin width is fine
    \item for coarse data need bin widths $\omega \ge \alpha$ (show plot with white space gaps)
    \item for coarse data bin widths should be scalar multiple of $\alpha$ (show plot with strips)
    \item Does the distribution type matter? 
  \end{itemize}
\end{itemize}
}

@@

Investigation of the loss sources we might be tempted to conclude that the standard binning algorithm is superior due to lower loss, however the random binning algorithm displays strong advantages when encountering problematic data structure issues. Random binning is especially advantageous when discrete data values are recorded at a consistent increments but we bin our data using a bin width that is a non-integer multiple of the incremental widths. In this scenario, we see that the non-synchronous data increments and bin widths will cause bins to contain inconstant numbers of possible data value. Assuming that data values are spread uniformly over the discrete values we want a visualization that displays uniform frequency. Using the standard binning algorithm the misalignment of data increment and bin widths will lead to and artificial frequency discrepancy that manifests in a striped pattern in the visualization. The random binning algorithm is expected to display the intended uniform frequency pattern because the randomization of data values to bin centers mitigates the problematic bin width selection.

@@

Another thing to note about Figure ~\ref{fig:LossComparison1} is that under the standard binning algorithm the numerical loss is a monotone increasing function of the bin width, but visual loss is not a strictly increasing function of bin width.  This is apparent in the order of colored lines based on the Strike Out binning size.  This is because the numerical loss function sums loss over all observations, which means the distance from the points to the numerical center of bins can only increase as bin width increases. Whereas the visual loss function sums loss over bin, thus as the bin width increases the number of bins decreases which leads to the potential that the overall visual loss may decrease even though the distance from numerical centers to visual centers may increase for individual bins.

@@

We may also use this example to highlight the advantages of random binning. The two binned scatterplots at the top of Figure~\ref{fig:binning-comparison} show bins of the same width, but clear differences between the binning results are evident. The standard binning algorithm has rounding artifacts that are clearly visible in the vertical strips. These vertical stripes are from the computational rounding issues encountered if we use the standard binning when we have the data values were recorded on an evenly spaces discrete scale (in our case, integers) and the bin widths are integer multiples of the data increments.  This reinforces the argument that in practice the random algorithm can produce a graph that better represents the true shape of the data than the standard algorithm even if the numerical and visual loss is higher.

\textbf{Are we sure that this is because of an odd bin widths and not just that the bins were not centered on an integer?} In contrast, when the bin widths are odd, the two binning algorithms yield nearly indistinguishable binned scatterplots, as shown in the bottom row of Figure~\ref{fig:binning-comparison}.\textbf{I think that strips disappearing in (5,25) bins is because of where we START the binning not that the bin widths are even}
